Expert-In-The-Loop Causal Discovery

### 2024-08-06 paper/main.tex
updates

Understanding the cause-and-effect relationships between variables is a
fundamental goal in many scientific questions as it reveals the underlying
mechanisms of observed phenomena and informs effective interventions or policy
decisions. \emph{Causal Discovery} methods aim to uncover these underlying
cause-effect relationships among random variables from observational data. The
problem of causal discovery has been explored in both the Directed Acyclic
Graphs (DAGs) and Structural Equation Models (SEMs) literatures, each adopting
a different approach. In the DAG literature, the focus has primarily been on
developing automated algorithms that can learn the causal structure from
datasets. In contrast, the SEM literature has focused more towards developing
tools that assist researchers in manually constructing or modifying models.

As a result of these efforts, numerous algorithms for causal discovery have
been proposed in the DAG literature, such as constraint-based methods like the
PC algorithm \citep{Spirtes2001} and Fast Causal Inference (FCI)
\citep{Spirtes2000}), score-based methods like Hill-Climb Search and Greedy
Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. However, the adaption of these algorithms in applied fields
remain limited, and researchers typically prefer to construct DAGs manually
based on their domain expertise \citep{Tennant2020, Petersen2021}. We believe this 
is the case because of problems with these algorithms when applied in practical
settings:

@@

To deal with both this issues, in this paper, we develop a measure of partial
association which can be using in combination of the CI test to rank the
violations in these CI tests. On the other hand, in Structural Equation Models
(SEMs) literature, more emphasis has been placed on developing tools to help
researchers manually construct models while integrating their domain expertise.
Researchers typically develop an initial model based on domain knowledge and
use various tools to guide modifications to improve the fit of their model to
data. This model construction process is known as \emph{Specification Search}.
A common method for specification search is using \emph{modification indices},
which identify potential changes to the model structure that improve the model
fit. These potential changes are then ranked by the ones that improve the fit
the most. This ranking allows the researchers to focus first on the suggestions
that can improve the model the most and while integrating their expertise to
choose the most optimal modifications.

@@

			\draw[->]  (age) to[bend right=20] (ed);
			\draw[->]  (sex) to (ed);
			\draw[->]  (age) to (hrpw);
			\draw[->]  (ed) to (hrpw);
			\draw[->]  (sex) to (hrpw);
			\draw[->]  (ed) to (income);
			\draw[->]  (hrpw) to (income);
			\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
			\draw[->]  (sex) to (income);
			\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
			\draw[->]  (sex) to (occ);	
		\end{scope}
		\begin{scope}[xshift=-3cm]
			\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
				\begin{minipage}{.2\textwidth}
					\begin{equation*}
						\begin{split}
							\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
							\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\textnormal{Age} &\ci \textnormal{Sex} \\
							\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
						\end{split}
					\end{equation*}
				\end{minipage}
				};
		\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Finish figure. Show an example of model testing using CIs}}
		\label{fig:ci_testing}
\end{figure}
\begin{figure}
	\caption{\todo[inline]{Figure showing an example of modification indices}}
\end{figure}

@@

% Add that we need a p-value as well.

@@

The value of this partial measure of association can be interpreted as - given
the current structure of the model how much of the observed correlation between
the variables $ X $ and $ Y $ in data is not explained by the model. To modify
the model based on this, we can look at the pair of variables that have an edge
between them. If any of these pair have a p-value greater than partial
association below our chosen threshold, we can remove that edge as it is
redundant. In the case when there are

@@

@@

>>>>

Understanding the cause-and-effect relationships between variables is a
fundamental goal in many scientific questions as it reveals the underlying
mechanisms of observed phenomena and informs effective interventions or policy
decisions. \emph{Causal Discovery} methods aim to uncover these underlying
cause-effect relationships among random variables from observational data. The
problem of causal discovery has been explored in both the Directed Acyclic
Graphs (DAGs) and Structural Equation Models (SEMs) literatures, each adopting
a different approach. In the DAG literature, the focus has primarily been on
developing automated algorithms that can learn the causal structure from
datasets. As a result of these efforts, numerous algorithms for causal
discovery have been proposed in the DAG literature, such as constraint-based
methods like the PC algorithm \citep{Spirtes2001} and Fast Causal Inference
(FCI) \citep{Spirtes2000}), score-based methods like Hill-Climb Search and
Greedy Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. In contrast, the SEM literature has focused more towards
developing tools that can assist researchers in manually constructing or
modifying models by allowing them to integrate their expert knowledge into the
model building process. Researchers typically develop an initial model based on
domain knowledge and use various tools to guide modifications to improve the
fit of their model to data. This model construction process is known as
\emph{Specification Search}. A common method for specification search is using
\emph{modification indices}, which identify potential changes to the model
structure that improve the model fit. These potential changes are then ranked
by the ones that improve the fit the most. This ranking allows the researchers
to focus first on the suggestions that can improve the model the most and while
integrating their expertise to choose the most optimal modifications.

Even though so many automated algorithms have been proposed, the adaption of
these algorithms in applied fields remain limited, and researchers typically
prefer to construct DAGs manually based on their domain expertise
\citep{Tennant2020, Petersen2021}. We believe this is the case because of
problems with these algorithms when applied in practical settings:

@@

To deal with both this issues, in this paper, we develop a measure of partial
association which can be using in combination of the CI test to rank the
violations in these CI tests. On the other hand, in Structural Equation Models
(SEMs) literature, more emphasis has been placed on developing tools to help
researchers manually construct models while integrating their domain expertise.

@@

			\draw[->]  (age) to[bend right=20] (ed);
			\draw[->]  (sex) to (ed);
			\draw[->]  (age) to (hrpw);
			\draw[->]  (ed) to (hrpw);
			\draw[->]  (sex) to (hrpw);
			\draw[->]  (ed) to (income);
			\draw[->]  (hrpw) to (income);
			\draw[->]  (occ) to[out=300, in=0, looseness=1.4] (income.east);
			\draw[->]  (sex) to (income);
			\draw[->]  (ed) to[out=90, in=150, looseness=1.3] (occ);
			\draw[->]  (sex) to (occ);	
		\end{scope}
		\begin{scope}[xshift=-3cm]
			\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
				\begin{minipage}{.2\textwidth}
					\begin{equation*}
						\begin{split}
							\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
							\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
							\textnormal{Age} &\ci \textnormal{Sex} \\
							\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
						\end{split}
					\end{equation*}
				\end{minipage}
				};
		\end{scope}
		\end{tikzpicture}
		\caption{\todo[inline]{Finish figure. Show an example of model learnt using PC algorithm $\rightarrow$ Orienting the unoriented edges $\rightarrow$  Model testing}}
		\label{fig:ci_testing}
\end{figure}
\begin{figure}
	\caption{\todo[inline]{Take a model; show the modification indices for it.}}
\end{figure}

@@

@@

For a pair of variables $ X $ and $ Y $, if there is an edge present between
them, the interpretation of this partial association measure is similar to path
coefficients in SEMs, i.e. how strong is the edge between the variables. Hence,
we can use this to identify potential edges that can be removed from the model.
In the case when there is no edge between the variables, the value of this
partial measure of association can be interpreted as - given the current
structure of the model how much of the observed correlation between the
variables $ X $ and $ Y $ in data is not explained by the model. To modify the
model based on this, we can look at the pair of variables that have an edge
between them. If any of these pair have a p-value greater than partial
association below our chosen threshold, we can remove that edge as it is
redundant.

The measure of association also gives us a way to rank the potential
modifications. The smallest values (below the threshold) of the association
measure are the worst offending ones and for when the edges are not present the
largest values would add the most amount of explainability to the model. Using
this ranking the user can decide which pair of variables to focus on.

@@

\todo[inline]{Add a figure showing an example of how this modification works.}

@@

\section{Connection to score based methods}

\begin{itemize}
	\item Because they make only local changes, can get stuck in local minimas.
\end{itemize}
